---
title: "Graphics Depicting Annual Frequencies of Extreme  Weather Events"
output: html_notebook
---

# Introduction
This R Notebook contains code producing graphics used in CBEP's 2020 State of the Bay report to depict long-term changes in the frequency of extreme weather events.  These are final revisions of earlier graphic drafts, so code and commentary are more limited.

# Install Libraries
```{r libraries}
library(tidyverse)
library(readr)

library(CBEPgraphics)

# And load the CBEP fonts too.
load_cbep_fonts()
```

# LOad Data
We  set up folder paths to our data, and access the data.  These data contain both temperature exceedences and precipitation exceedences.
```{r data_access}
sibfldnm <- "Original Data"
parent <- dirname(getwd())
sibling <- file.path(parent, sibfldnm)
fn <- "longannualdata.csv"

longannualdata <- read_csv(file.path(sibling, fn)) %>%
  select(-station) %>%
  mutate(year = as.numeric(format(date, format = "%Y")))
```

# Temperature Exeedences
The Annual Data Downloaded from NOAA's Climate Data Online includes counts of the number of days exceeding various temperature thresholds, as follows:  

* DX90: Number of days with maximum temperature >= 32.2°C/90°F.
* DX70: Number of days with maximum temperature >= 21.1°C/70°F.
* DX32: Number of days with maximum temperature <= 0°C/32°F.
* DT32: Number of days with minimum temperature <= 0°C/32°F.
* DT00: Number of days with minimum temperature <= -17.8°C/0°F.

We focus on days that got over over 90 Fahrenheit, days that dropped below 32, and days that got as low as 0 degrees F. Notice that even though the CDO data was downloaded in metric units, the cumulative counts are based on English units.

## Data Reorganization
```{r temp_data_organization}
textdatalong <- longannualdata %>%
  filter(datatype %in% c('DX90', 'DT32', 'DT00')) %>%
  mutate(datatype = factor(datatype, levels = c('DT00', 'DX90', 'DT32')))

textdata <- longannualdata %>%
  filter(datatype %in% c('DX90', 'DT32', 'DT00')) %>%
  spread(key = datatype, value = value)
```

## Assemble Plot
This is a complex graphic that rests on Poisson GLM models to offer predicted values which turn up in annotations later.  To assemble the plot, we first need to run the Poisson GLMs, then extract predictions and finally build up a complex plot.

Where these are relatively simple GLMs, we can show prediction lines using geom_smooth(), but here we also use the GLM models to provide decadal predictions as "typical" values from an era.

### Poisson GLM Models
Note that while I prefer the quasi-Poisson models for inference, the predictions of the the Poisson models are identical. They differ only in dispersion.  For plotting purposes, since I don't show model dispersion on the plots, the two are functionally identical.

```{r temp_models}
gt90GLM <- glm(DX90 ~ year, family = poisson, data = textdata)
lt32GLM <- glm(DT32 ~ year, family = poisson, data = textdata)
lt00GLM <- glm(DT00 ~ year, family = poisson, data = textdata)
```

### Decadal Predictions of days with warm or cold weather
We want predictions to offer "typical" values from the 1940s and the 2010s in subsequent plots.  We look at predictions from the middle of each decade, as generally representative of that era.
```{r temp_decadal_predictions}
newdf <- data.frame(year = c(1945, 1955, 1965, 1975, 1985, 1995, 2005, 2015))
gt90pred <- predict(gt90GLM, newdata = newdf, type = "response")
lt32pred <- predict(lt32GLM, newdata = newdf, type = "response")
lt00pred <- predict(lt00GLM, newdata = newdf, type = "response")

predict_df <- tibble(year = newdf$year,
                     gt90 = gt90pred,
                     lt32 = lt32pred,
                     lt00 = lt00pred)
rm(newdf)
knitr::kable(predict_df, digits = 0)
```

### Construct Annotations Tibble
```{r}
labs<-tibble(year=rep(c(1940, 1995),3),
             y = c(45,30,130, 120, 25,20),
             txt= c(paste('~', round(predict_df$lt00[1],0), 'per year'),
                    paste('~', round(predict_df$lt00[8],0), 'per year'),
                    paste('~', round(predict_df$lt32[1],0), 'per year'),
                    paste('~', round(predict_df$lt32[8],0), 'per year'),
                    paste('~', round(predict_df$gt90[1],0), 'per year'),
                    paste('~', round(predict_df$gt90[8],0), 'per year')),
             datatype = rep(c('DT00', 'DT32', 'DX90'), each=2)
             )
```

### Create Facet Headers
```{r temp_facet_headers}
type.labs <- c('Below Zero', 'Below Freezing', 'Above 90 F')
names(type.labs) <- c('DT00',  'DT32', 'DX90')
```

### Single Facet Plot
```{r t_exceedences_plot, fig.width = 9, fig.height =4, warning=FALSE}
plt <- ggplot(textdatalong, aes(x = year, y = value, color = datatype)) +
  geom_point() +
  geom_text(aes(x = year, y = y, label = txt),
            color = "black", data = labs,
            hjust = 0, size = 3) +
  geom_smooth(aes(linetype = datatype), method = glm, formula = y ~ x,
              method.args = list(family = "poisson"),
              se = FALSE) + 
  
  theme_cbep(base_size = 14) + 
  theme(legend.position = "None") + 
  theme(axis.text = element_text()) + 
  
  xlab("Year") + ylab("Days per Year") + 
  
  scale_color_manual(values = cbep_colors2(), name = "",
                     labels = c("Below Zero",
                                parse(text = "Above~90*degree*F"),
                                "Below Freezing")) +
  scale_linetype_manual(values = c("solid", "blank", "solid"))


plt +
  facet_wrap(~datatype, labeller = labeller(datatype = type.labs)) +
  theme(panel.spacing = unit(2.5, "lines"))

ggsave("frequencythreehoriz.pdf", device = cairo_pdf, width = 9, height = 4)
ggsave("frequencythreehoriz.png", width = 9, height = 4, type = "cairo")
```

### Three Separate Small Plots
We use a loop to generate separate plots that mimic the combined plots just prepared.
```{r three_plots, fig.width = 5, fig.height = 3}
# we can reuse type.labs here, slightly modified to allow parsing
type.labs <- c('Below Zero', 'Below Freezing', parse(text = 'Above~90*degree*F'))
names(type.labs) <- c('DT00', 'DT32', 'DX90')
mycolors <- cbep_colors()[1:3]
names(mycolors) <- names(type.labs)
```

```{r revised_temp_annotations}
labs$year<- rep(c(1940, 2000),3)
labs$y   <- c(40,25,135, 125, 20,15)
```

```{r, fig.width = 5, fig.height = 3}
for (param in levels(textdatalong$datatype)) {
  tmp <- textdatalong %>% filter(datatype == param)
  annot <- labs %>% filter(datatype == param)
  
  plt <- ggplot(tmp, aes(x = year, y = value)) +
    geom_point(color = mycolors[param]) +
    geom_text(aes(x = year, y = y, label = txt), data = annot, hjust = 0, size = 4 ) +
    geom_smooth(method = glm, formula = y~x,
                  method.args = list(family = "poisson") ,
                  se= FALSE, color=mycolors[param]) +
      
    theme_cbep(base_size = 14) +
    theme(legend.position = "None") +
    theme(axis.text = element_text()) +
    
    xlab('Year') +
    ylab("Days per Year") +
    ggtitle(type.labs[param])
    
  print(plt)
  
  fnroot <- c('Below_Zero', 'Below_Freezing', 'Above_90_F')
  names(fnroot) <- c('DT00',  'DT32', 'DX90')
  fn=paste0(fnroot[param], ".pdf")
  ggsave(fn, device = cairo_pdf, width = 5, height = 3)
  fn=paste0(fnroot[param], ".png")
  ggsave(fn, width = 5, height = 3, type = 'cairo')
}
rm(tmp, annot, fnroot)

```

```{r temp_cleanup, echo=FALSE}
rm(lt00GLM, lt32GLM, gt90GLM, labs, textdata, textdatalong)
rm(plt, gt90pred, lt00pred, lt32pred, type.labs, labs, mycolors)
```

# Precipitation Frequencies
## Data Reorganization
The Annual Data Downloaded from NOAA's Climate Data Online includes counts of the number of days exceeding various precipitation thresholds, as follows:
* DP01: Number of days with >= 0.01 inch/0.254 millimeter in the month (year).
* DP10: Number of days with >= 0.1 inch/2.54 millimeter in the month (year).
* DP1X: Number of days with >= 1.0 inch (25.4mm) precipitation in the month (year).
* DSNW: Number of days with snowfall >= 1 inch (25 mm).
* DSND: Number of days with snow depth >= 1 inch (25 mm).

Note this does not include the number of days with more than two inches of rain, which is arguably a better metric of intense storms than the one inch rain threshold.  We will derive that metric directly from the daily weather data.

```{r precip_data_reorganization}
pextdata <- longannualdata %>%
  filter(datatype %in% c('DP01', 'DP1X', 'DSNW', 'DSND')) %>%
  select(-attributes) %>%
  spread(key = datatype, value = value)
```

###  Calculate Annual Days Exceeding Two Inches Rainfall
We first access the daily data, and calculate the number of days greater than or equal to two inches of rainfall.  We then join those data back into our database of precipitation exceedences, making sure not to drop years with zero days with more than two inches of rain.  
```{r daily_two_inch_exceedences}
sibfldnm <- 'Original Data'
parent <- dirname(getwd())
sibling <- paste(parent, sibfldnm, sep = '/')
fn <- 'longdailydata.csv'

num2inch <- 
  # Read daily data
  read_csv(file.path(sibling, fn)) %>%
  select(-station) %>%
  mutate(year = as.numeric(format(date, format='%Y')))  %>%
  select(date, year, datatype, value) %>%
  
  # Filter down to only precipitation data
  filter(datatype == 'PRCP') %>%
  rename(PRCP = value) %>%
  filter( ! is.na(PRCP)) %>%
  
  # Convert to inches. Original data in tenths of mm
  mutate(PRCP_IN = PRCP * 0.0393700787 / 10) %>% 
  filter(year > 1940) %>%   # 1940 is an incomplete year in the daily data
  
  # And count the number of high rain events
  filter(PRCP_IN >= 2) %>%  # Only days with two inches of rain or more
  group_by(year) %>%
  summarise(COUNT = n())    # use all caps because "count" is a bultin function
```

### Combined Data
The method just used to count the number of days with more than two inches of rain dropped any years in which there were no such days.  We want years with zero intense storms to be included in the data, only with the value of zero.  One can accomplish that using a left_join back into a data frame that includes all years in the historical data record for the Portland Jetport.
```{r final_precip_data}
pextdata <- left_join(pextdata, num2inch, by = 'year') %>%
  select(-DP01, -DSND, -DSNW) %>%
  mutate(COUNT=replace_na(COUNT,0)) %>%
  rename(DP2X = COUNT)

pextdatalong <- pextdata %>%
  pivot_longer( -c(date,year), names_to='datatype', values_to = 'value')

rm(sibfldnm, parent, sibling, fn, num2inch)
```

## Assemble Plot
### Poisson GLM Models
Again, since we only use these to generate predictions, we don't need the wider uncertainty intervals from the quasi-Poisson GLM models.
```{r precip_glm_models]}
gt1GLM <- glm(DP1X ~ year, family = poisson, data = pextdata)
gt2GLM <- glm(DP2X ~ year, family = poisson, data = pextdata)
```

### Decadal Predictions
```{r precip_predictions}
newdf <- tibble(year=c(1945, 1955, 1965, 1975, 1985, 1995, 2005, 2015))
predict_df <- tibble(year = newdf$year,
                     gt1  = predict(gt1GLM, newdata=newdf, type='response'),
                     gt2  = predict(gt2GLM, newdata=newdf, type='response'))
knitr::kable(predict_df, digits=0)
```

So, days with more than one inch of rain have gone from about eight a year in the 1940s to about thirteen a year in the 2010s and days with two or more inches of rain have gone from about one a year to about three a year.

### Construct Annotations Tibble
```{r precip_annotation_tibble}
labs<-tibble(x =c(1940, 2000, 1940, 2000),
             y = c(18, 22, 4, 7),
             txt= c(paste('~', round(predict_df$gt1[1],0), 'per year'),
                    paste('~', round(predict_df$gt1[8],0), 'per year'),
                    paste('~', round(predict_df$gt2[1],0), 'per year'),
                    paste('~', round(predict_df$gt2[8],0), 'per year')),
             datatype = rep(c('DP1X', 'DP2X'), each = 2))
```

### Draw Basic Plot
```{r precip_exceedences_plot, warning=FALSE}
plt <- pextdatalong %>%
  filter(datatype %in% c('DP1X', 'DP2X')) %>%
  
  ggplot(aes(x = year, y = value))+
  geom_point(aes(color = datatype)) +
  geom_smooth(mapping = aes(color = datatype), method = glm, formula = y~x,
              method.args = list(family = "poisson") ,
              se = FALSE) +
  
  geom_text(data = labs,
            mapping = aes(x = x, y = y, label = txt, group = datatype),
            hjust = 0, size = 3) +
  
  xlab('Year') +
  ylab("Days per Year") +
  scale_color_manual(values = cbep_colors(), name = '',
                      labels = c('> 1 inch Precip.', 
                                 '>2 inch Precip.')) +
  theme_cbep()

plt +  theme(legend.position = c(.15, .9)) +
  guides(color = guide_legend(override.aes = list(linetype = 0)))
```

### Draw Faceted Plot
```{r precip_exceedences_plot_facets, fig.width = 8, fig.height = 4, warning = FALSE}
plt + facet_wrap(~datatype, nrow = 1) +
  theme_cbep(base_size = 12) +
  theme(strip.text.x = element_blank(),
        legend.position = c(.85, .9),
        panel.spacing = unit(2, "lines")) +
   guides(color = guide_legend(override.aes = list(linetype = 0)))

ggsave('highrainfalldays.png', width = 8, height = 4,  type = 'cairo')
ggsave('highrainfalldays.pdf', device = cairo_pdf, width = 8, height = 4)
```

### Draw Alternate Faceted Plot
```{r precip_exceedences_plot_facets_2, fig.width = 8, fig.height = 4, warning = FALSE}
plt + facet_wrap(~datatype, nrow = 1, scales = 'free_y') +
  theme_cbep(base_size = 12) +
  theme(strip.text.x = element_blank(),
        legend.position = c(.65, .95),
        panel.spacing = unit(2, "lines")) +
   guides(color = guide_legend(override.aes = list(linetype = 0)))

ggsave('highrainfalldays_free.png', width = 8, height = 4,  type = 'cairo')
ggsave('highrainfalldays_free.pdf', device = cairo_pdf, width = 8, height = 4)
```

### Draw Single Plot of Precip. > 1 in.
```{r precip_exceed_1_plot, fig.width = 5, fig.height = 3, warning = FALSE}
tmp <- labs[labs$datatype == 'DP1X',]

plt <- pextdata %>%
  ggplot(aes(x = year, y = DP1X))+
  geom_point(color = cbep_colors()[1]) +
  
  geom_smooth(method = glm, formula = y~x,
              method.args = list(family = "poisson") ,
              se = FALSE, color = cbep_colors()[1]) +
  
  geom_text(mapping = aes(x = x, y = y, label = txt, group = datatype),
            data = tmp, hjust = 0) +
  
  scale_y_continuous(limits = c(0,25)) +
  
  xlab('Year') +
  ylab("Days > 1 in. of Rain") +
  theme_cbep(base_size = 12) 
plt
ggsave('days_gt1.png', width = 5, height = 3,  type = 'cairo')
ggsave('days_gt1.pdf', device = cairo_pdf, width = 5, height = 3)
```

### Draw Single Plot of Precip. > 2 in.
```{r precip_exceed_2_plot, fig.width = 5, fig.height = 3, warning = FALSE}
tmp <- labs[labs$datatype == 'DP2X',]

plt <- pextdata %>%
  ggplot(aes(x = year, y = DP2X))+
  geom_point(color = cbep_colors()[2]) +
  
  geom_smooth(method = glm, formula = y~x,
              method.args = list(family = "poisson") ,
              se= FALSE, color=cbep_colors()[2]) +
  
  geom_text(mapping = aes(x = x, y = y, label = txt, group = datatype),
            data = tmp, hjust = 0) +
  
  scale_y_continuous(limits=c(0,8)) +
  
  xlab('Year') +
  ylab("Days > 2 in. of Rain") +
  theme_cbep(base_size = 12) 
plt
ggsave('days_gt2.png', width = 5, height = 3,  type = 'cairo')
ggsave('days_gt2.pdf', device = cairo_pdf, width = 5, height = 3)
```
