---
title: "Analysis of Portland Jetport Historical Weather Data"
output: html_notebook
---

# Install Libraries
```{r}
library(tidyverse)
library(readr)
library(ggthemes)
library(extrafont)
library(svglite)
#library(CBEPgraphics)
```

# Set Graphic Defaults
This code block calls a script that establishes defaults for a consistent Casco Bay State of the Bay graphics look and feel.  It establishes cbep_colors, cbep_colors2, and theme_cbep, after loading selected fonts.  It then sets theme_cbep() as the default theme.

If you do not run this code, you will need to define the CBEP color names somewhere else, or remove references to them in the code.
```{r}
fn <- 'CBEPGraphicsSetup.R'
source(fn)
rm(fn)
```

# Read Data
```{r}
sibfldnm <- 'Original Data'
parent <- dirname(getwd())
sibling <- paste(parent,sibfldnm, sep = '/')
fn <- 'longannualdata.csv'

longannualdata <- read_csv(paste(sibling,fn, sep = '/')) %>%
  select(-station) %>%
  mutate(year = as.numeric(format(date, format='%Y'))) %>%
  mutate(cyear = year-1980)
```

# Annual Minimum, Maximum and Average Temperature
Codes for Key Temperature records in the downloaded annual data are:
TAVG  -- Annual Average Temperature (unweighted; effectively average of monthly averages, of daily maxes and mins.  Note that this means this is not indpendent of the other two metrics.)
EMXT  -- Extreme Maximum Annual Temperature
EMNT  -- Extreme Minimum Temperature

```{r}
tdata <- longannualdata %>%
  filter(datatype %in% c('TAVG', 'EMXT', 'EMNT')) %>%
  mutate(value = value*(9/5)+32) %>%
  mutate(datatype = factor(datatype, levels = c('EMXT', 'TAVG', 'EMNT' )))
```

## Use lm to calculate slopes
But since I'm using just linear models here, I can use geom_smooth() provided I'm OK with plotting all three regression lines
```{r}
tdataF <- longannualdata %>%
  filter(datatype %in% c('TAVG', 'EMXT', 'EMNT')) %>%
  mutate(value = value*(9/5)+32) %>%
  mutate(datatype = factor(datatype, levels = c('EMXT', 'TAVG', 'EMNT' ))) %>%
  spread(key = datatype, value = value)
mean.lm <- lm(TAVG~ cyear, data = tdataF)
min.lm <- lm(EMNT~ cyear, data = tdataF)
# max.lm <- lm(EMXT~ cyear, data = tdataF) NS trend....
```

```{r}
plt <- ggplot(tdata, aes(year, value, color=datatype)) + geom_line() +
  xlab('Year') +
  ylab (expression('Temperature ('*~degree*F*')')) +
  scale_color_manual(values =cbep_colors2, name = 'Annual Values',
                     labels = c('Maximum', 'Average', 'Minimum'))
plt
```


```{r width = 7, height = 5}
plt + # geom_smooth(method= 'lm', se=FALSE, linetype=3)
  geom_abline(slope = mean.lm$coef[2], intercept = mean.lm$coef[1] - 1980 *mean.lm$coef[2],
              lty = 3) +
  geom_abline(slope = min.lm$coef[2], intercept = min.lm$coef[1] - 1980 *min.lm$coef[2],
              lty = 3)

```
That graphic is not very satisfying, as the slopes are difficult to perceive.  I wonder if it might be better to present these as temperature anomolies, either from a designated year or from the long term averages.


```{r}
tdata.centered <- longannualdata %>%
  filter(datatype %in% c('TAVG', 'EMXT', 'EMNT')) %>%
  mutate(value = value*(9/5)+32) %>%
  select(-attributes) %>%
  spread(key = datatype, value = value) %>%
  mutate_at(c('TAVG', 'EMXT', 'EMNT'), ~ scale(., scale = FALSE)) %>%
  gather(key = 'datatype', value = 'value', -date, -year, -cyear) %>%
  mutate(datatype = factor(datatype, levels = c('EMXT', 'TAVG', 'EMNT' ))) 

tdata.centered.wide <- tdata.centered %>%
  spread(key = datatype, value = value)

mean.lm.2 <- lm(TAVG~ cyear, data = tdata.centered.wide)
min.lm.2 <- lm(EMNT~ cyear, data = tdata.centered.wide)
max.lm.2 <- lm(EMXT~ cyear, data = tdata.centered.wide)
```

Can also fill in anomolies with different colors, as follows.
```{r}
type.labs <- c('Maximum', 'Average', 'Minimum')
names(type.labs) <- c("EMXT", "TAVG", "EMNT")

plt1 <- ggplot(tdata.centered, aes(year, value)) + 
  geom_bar(aes(fill = ifelse(value <0, "red4", "blue")),stat = "identity",
                                                                 position = "identity") +
  scale_fill_discrete(name = '', labels = c('Above Average', 'Below Average')) +
  xlab('Year') +
  ylab (expression('Temperature Anomaly ('*~degree*F*')')) +
  facet_wrap('datatype', nrow=3, scales = 'fixed', # alternative is "free_y"
             labeller = labeller(datatype = type.labs),
             strip.position = 'right') +
  theme(legend.position = "none")
plt1
```

## Create Lines to plot
Note the way I assemble a mathematical expression to be interpreted by "plotmath" for the annotations.  The key idea is that you need to link each symbol or word either with "~" if you want a space, or "*" if you don't.
```{r}

lines <- tibble(datatype = c("EMXT", "TAVG", "EMNT"),
               theslope = c(max.lm.2$coef[2], mean.lm.2$coef[2], min.lm.2$coef[2]),
               theintercept = c(max.lm.2$coef[1] - 1980 *max.lm.2$coef[2],
                                mean.lm.2$coef[1] - 1980 *mean.lm.2$coef[2],
                                min.lm.2$coef[1] - 1980 *min.lm.2$coef[2])) %>%
    mutate(datatype = factor(datatype, levels = c('EMXT', 'TAVG', 'EMNT' ))) 

labs <-  tibble(datatype = c("EMXT", "TAVG", "EMNT"),
              ypos1 = c(-8, -8, -8),
              ypos2 = c(-2, -1.5, -5),
              xpos = c(1950, 2000, 2000),
              txt = c("No~trend",
                       paste(round(mean.lm.2$coef[2]*10,2), '~degree*F', "~per~decade"),
                       paste(round(min.lm.2$coef[2]*10,2), '~degree*F', "~per~decade")) ) %>%
  mutate(datatype = factor(datatype, levels = c('EMXT', 'TAVG', 'EMNT' ))) 
```



```{r fig.width = 7, fig.height = 5}
plt1 +
  geom_abline(aes(slope = theslope, intercept = theintercept), data = lines, lty = 1, alpha = 0.75) +
  geom_text(aes(x=xpos, y=ypos1, label= txt), data = labs, parse=TRUE)

ggsave('tempanomolies1.png', width=7, height=5,  type = 'cairo')
#ggsave('tempanomolies1.svg', width=7, height=5)
ggsave('tempanomolies1.pdf',
       device=cairo_pdf, width = 7, height = 5)
```

# Precipitation Totals
Codes for key annual PRECIPITATION records in the downloaded data are:
PRCP:  Total Monthly (Annual) precipitation. Precipitation totals are based on daily or multi-day (if daily is missing) precipitation report, in millimeters to tenths.
SNOW:  Total Monthly (Annual) Snowfall in millimeters.
## Read Data and Transform to English Units.
```{r}
pdatalong <- longannualdata %>%
  filter(datatype %in% c('PRCP', 'SNOW')) %>%
  mutate(datatype = factor(datatype)) %>%
  mutate(value = value*0.0394)           # Convert to inches from tenths of mm

pdata <- longannualdata %>%
  filter(datatype %in% c('PRCP', 'SNOW')) %>%
  mutate(datatype = factor(datatype)) %>%
  mutate(value = value*0.0394) %>%
  spread(key=datatype, value=value)
```

# Initial Plot
```{r fig.width = 7, fig.height = 5}
plt <- ggplot(pdatalong, aes(year, value, color=datatype)) + 
  geom_point(size=2) +
  xlab('Year') +
  ylab('Precipitation (inches)') +
  scale_color_manual(values=cbep_colors, name = '', labels = c('Precipitation', 'Snow')) +
  ylim(c(0,150))
plt + geom_smooth(method='lm', se=FALSE, lty=1) 
```
They don't work well on one plot.
```{r}
snow.lm <- lm(SNOW~year, data = pdata)
precip.lm <- lm(PRCP~year, data = pdata)
```

```{r fig.width = 7, fig.height = 5}
lines <- tibble(yr=seq(1941,2019), prcp= precip.lm$coef[1] + precip.lm$coef[2]*seq(1941,2019))
plt + geom_line(data = lines, aes(x=yr, y=prcp), lty=1, color=cbep_colors[1])
#ggsave('draftprecipsnow.png', width = 7, height = 5,  type = 'cairo')
#ggsave('draftprecipsnow.svg', width = 7, height = 5)
#ggsave('draftprecipsnow.pdf', device=cairo_pdf, width = 7, height = 5)
```

```{r fig.width = 7, fig.height = 5}
plt <- ggplot(pdata, aes(year, PRCP)) + 
  geom_point(size=2) +
  geom_smooth(method='lm', se=FALSE, lty=1) +
  xlab('Year') +
  ylab('Precipitation (inches)') +
  ylim(c(0,80)) 
plt
```

```{r fig.width = 7, fig.heigth = 5}
txt <- paste('+~', round(precip.lm$coef[2]*10,2), '~inches', "~per~decade")
plt + 
  geom_text(aes(x=1995, y=25, label= txt), parse=TRUE, hjust=0)

ggsave('precip.png', width = 7, height = 5, type = 'cairo')
#ggsave('precip.svg', width = 7, height = 5)
ggsave('precip.pdf', device=cairo_pdf, width = 7, height = 5)
```



