---
title: "Analysis of Length of Growing Season"
output: html_notebook
---

# Load Libraries
```{r}
library(tidyverse)
library(readr)
library(mblm)
```

# Set Graphic Defaults
This code block calls a script that establishes defaults for a consistent Casco Bay State of the Bay graphics look and feel.  It establishes cbep_colors, cbep_colors2, and theme_cbep, after loading selected fonts.  IF you do not run this code, you will need to define those names somewhere else, or remove references to them in the remainder of the code.
```{r}
sibfldnm <- 'Graphics'
parent <- dirname(getwd())
sibling <- paste(parent,sibfldnm, sep = '/')
fn <- 'CBEPGraphicsSetup.R'

source(paste(sibling,fn, sep = '/'))
rm(sibfldnm, parent, sibling, fn)
```

# Read Data
The frost-free period needs to be derived directly from the daily temperature data.  Note that temperature in the source data is in tenths of a degree C.
```{r}
sibfldnm <- 'Data'
parent <- dirname(getwd())
sibling <- paste(parent,sibfldnm, sep = '/')
fn <- 'longdailydata.csv'

longdailydata <- read_csv(paste(sibling,fn, sep = '/')) %>%
  select(-station) %>%
  filter(datatype %in% c('TMAX', 'TMIN')) %>%
  mutate(doy =  as.numeric(format(date, format='%j'))) %>%
  mutate(month = as.numeric(format(date, format='%m'))) %>%
  mutate(year = as.numeric(format(date, format='%Y'))) %>%
  mutate(cyear = year-1980)
```

# Find First and Last Frost Dates
The following strategy was suggested by a Stack Overflow answer here:  
https://stackoverflow.com/questions/41328946/find-last-values-by-condition

In brief, we filter a dataframe down to data that meets our criteria -- here, days with temperatures below freezing -- and then group_by() to isolate data from each year, and finally slice() to pick the first or last item in the list.  Be aware that this strategy depends on the data being in chronological order.
```{r}
cutoffinF <- 32
cutoffinC <- (cutoffinF − 32) * 5/9

firstfrosts <- longdailydata %>%
  spread(key = datatype, value = value) %>%
  filter(year>1940) %>%                 # 1940 is an incomplete year in the daily data
  mutate(TMIN = TMIN/10) %>%            # raw data is in tenths of degree C
  filter(TMIN<= cutoffinC) %>%          # Only days below freezing
  filter(month>=7) %>%                  # only the last half of the year
  group_by(year) %>%
  slice(1)  %>%                         # Choose the first record
  select(date, year, doy)

lastfrosts <- longdailydata %>%
  spread(key = datatype, value = value) %>%
  filter(year>1940) %>%                 # 1940 is an incomplete year in the daily data
  mutate(TMIN = TMIN/10) %>%            # raw data is in tenths of degree C
  filter(TMIN<= cutoffinC) %>%          # Only days below freezing
  filter(month<7) %>%                   # only the first half of the year
  group_by(year) %>%
  slice(n())  %>%                       # Choose the last record
  select(date, year, doy)

frosts <- inner_join(firstfrosts,lastfrosts, by = "year", suffix = c('.first', '.last')) %>%
  mutate(frostfree = doy.first - doy.last - 1) # calculate frost-free period

rm(firstfrosts,lastfrosts)
```

# Statistical Analysis
## Frost Free Period
```{r}
ffdays.lm <- lm(frostfree~ year, data = frosts)
summary(ffdays.lm)
```
So, on average, the number of frost free days over the last 80 years has declined by about half a day a year, or about 39 days overall.
```{r}
 round(coef(ffdays.lm)[2]*80,2)
```
```{r}
oldpar <- par(mfrow=c(2,2))
plot(ffdays.lm)
par(oldpar)
```
Theoretically, we know we are dealing with count data, so perhaps something other than a Gaussian model (Poisson or quasi-Poisson GLM?)  would perform better, but we don't see any strong deviations (here) from normality. That's not unreasonable since the binomial distribution for high counts converges on a normal distribution.

Curvature in the residuals suggests perhaps the relationship is, non-linear.  We could fit either a piece-wise linear function or a quadratic term to explore possibilities. But a quadratic term dramatically degrades the quality of the model fit.
```{r}
ffdays.lm2 <- lm(frostfree~ year + I(year*year), data = frosts)
anova(ffdays.lm, ffdays.lm2)
```
So, no real evidence of non-linearities that matter.
## Date of First Frosts
```{r}
firstfrosts.lm <- lm(doy.first ~ year, data = frosts)
summary(firstfrosts.lm)
```
Slightly more than half of the trend is due to later first frosts, about 23 days.
```{r}
 round(coef(firstfrosts.lm )[2]*80,2)
```
```{r}
oldpar <- par(mfrow=c(2,2))
plot(firstfrosts.lm)
par(oldpar)
```

## Date of Last Spring Frost
```{r}
lastfrosts.lm <- lm(doy.last ~ year, data = frosts)
summary(lastfrosts.lm)
```

So last frosts have moved just over two weeks earlier over eighty years:
```{r}
round(coef(lastfrosts.lm)[2]*80,2)
```


# Robust alternatives to Gaussian Models
The corresponding Thiel-Sen estimators are similar.  Here I am using the 'mblm' or "Median Based Linear Models" functions from the mblm package.  This fits a Theil-Sen estimator, which is a more robust version of fitting a linear trend.  I also run a Kendal's Tau test, which is (roughly) a test that checks if the Theil-Sen estimator is statistically significant.
```{r}
ffdays.ts <- mblm(frostfree~ year, dataframe = frosts)
summary(ffdays.ts)
cor.test(~frostfree+year, data = frosts, method = 'kendall')
```
```{r}
firstfrosts.ts <- mblm(doy.first ~ year, dataframe = frosts)
summary(firstfrosts.ts)
cor.test(~doy.first+year, data = frosts, method = 'kendall')
```
```{r}
lastfrosts.ts <- mblm(doy.last ~ year, dataframe = frosts)
summary(lastfrosts.ts)
cor.test(~doy.last+year, data = frosts, method = 'kendall')
```


# Draft Graphics
Note that as all I fit in this context wer esimple linear models, I can create graphics using geom_smooth() with method ='lm'.  This just draws teh linear model.
```{r}
plt <- ggplot(frosts, aes(year, frostfree)) + geom_point() +
  geom_smooth(method = 'lm', se=FALSE, lty=2) +
  ylab('Frost Free Days') +
  xlab('Year') +
  theme_cbep()
plt
```

## Show all three indicators
```{r}
pltdat <- frosts %>%
  select(-date.first, -date.last) %>%
  gather(key = 'indicator', value = 'days', -year) %>%
  mutate(indicator = factor(indicator,
                            levels = c("frostfree", "doy.first", "doy.last"),
                            labels = c("Frost Free Days", "First Frost", "Last Frost")))
```

```{r}
plt <-ggplot(pltdat, aes(year, days, color=indicator)) + geom_line(lwd = 1) +
  geom_smooth(method = 'lm', se = FALSE, lty=2) +
  scale_color_manual(values=cbep_colors, name = '', labels = c('Frost Free Days', 'First Frost', 'Last Frost')) +
  theme_cbep() +
  xlab('Year') +
  ylab('Days')
plt

```
I find that confusing, as the "days" and "Day of Year" are both on the vertical axis, in a confusing way.
Lets split 
```{r}
plt +
  theme(legend.position = 'none') +
  facet_wrap('indicator', nrow=3, scales='free_y',
             strip.position = 'right') +
  ylab('Days / Day of Year') +
  scale_color_manual(values=cbep_colors, name = '', labels = c('Frost Free Days', 'First Frost', 'Last Frost')) +
  theme_cbep() +
  theme(panel.spacing = unit(.5, "lines")) +
  theme(legend.position = 'none')
```

# Final Graphic
I don't like the compound vertical axis.  A better solution is to use the strip labels as axis labels, as explained here:
https://stackoverflow.com/questions/37573155/showing-different-axis-labels-using-ggplot2-with-facet-wrap

I also increase the size of the graphic slightly, to provide room for the y axis labels.
```{r fig.width = 7, fig.height=5}
plt1 <- plt +
  ylab(NULL) +

  theme(legend.position = 'none') +
  facet_wrap('indicator', nrow=3, scales='free_y',
             strip.position = 'left') +
  theme(strip.background = element_blank(),
        strip.placement = "outside",
        strip.text = element_text(size = 12),
        panel.spacing = unit(0.5, "lines")) +
    theme(legend.position = 'none')
plt1
ggsave('frostfreedaysthreeaxes2.png', width = 7, height = 5, units='in')
```


#3 And output a PDF version without the Corbel Font
```{r}
plt2 <- plt1 + theme_cbep(base_family='sans')
ggsave('frostfreedaysthreeaxes2.pdf', plot=plt2, 
       device='pdf',
       width = 7, height = 5, units='in')
print(plt2)     # Plot 1 --> in the first page of PDF
dev.off() 

```

## Show the first and last frost dates only
```{r}
tmp <- frosts %>%
  select(-doy.first, -doy.last, -frostfree) %>%
  gather(key = 'indicator', value = 'date', -year) %>%
  mutate(month = format(date, format = '%m')) %>%
  mutate(day = format(date, format = '%d' )) %>%
  mutate(day= ifelse(day=='29' & month == '02', '28', day)) %>%   # Check for unlikely event of leap day
  mutate(compdate = as.Date(paste(month,day,'2019', sep = '/'), format = '%m/%d/%Y'))
```

```{r}
plt <- ggplot(tmp, aes(year, compdate, color=indicator)) + geom_line(lwd = 1) +
  geom_smooth(method = 'lm', se = FALSE, lty=2) +
  scale_color_manual(values = cbep_colors, name = '', labels = c('First Frost', 'Last Frost')) +
  theme_cbep() +
  xlab('Year') +
  ylab('Day of Year')

plt +
  scale_y_date(date_breaks='1 month', date_labels = '%b %e') +
  theme(legend.position=c(0.75,0.5))
```
